---
title: "DESeq2 Demo"
format: 
  html:
    page-layout: full
    grid:
      body-width: 1400px
      margin-width: 250px
      sidebar-width: 250px
    embed-resources: true
    toc: true
    toc-location: left
    toc-depth: 3
editor: 
  markdown: 
    wrap: 72
---

# Load libraries
= pak::pkg_install(<name>)
- pak::pkg_install("bioc::<library_name>")
```{r}
#| column: screen-inset
suppressPackageStartupMessages({
  library(DESeq2)
  library(tidyverse)
  library(clusterProfiler)
  library(biomaRt)
  library(fgsea)
  library(org.Hs.eg.db)
  library(patchwork)
})

SRC_DIR <- if (dir.exists("src")) "src" else "../src"
helper_file = file.path(SRC_DIR, "RNAseq_helpers.R")
source(helper_file)
```

# Setup the parameters. Hard-coded stuffs

```{r}
# gene types to work with
GENE_TYPE_PICK = c('protein_coding')

# padj threshold
PADJ_THRES = 0.05

# number of labels on the volcano
NLABEL = 50

# Max number of pathways in enrichGO barplots
ENRICH_BARCOUNT = 20

# for enrichGO
PVALUECUTOFF = 0.05

# Rank-proportional decrement used to break ties in the fgsea ranking metric
RANK_TIE_DELTA <- 1e-10

# signature genes
SIGNATURE_GENES <- c(
    "ESR1", "PGR", "ERBB2", "BRCA1", "BRCA2",
    "TP53", "MKI67", "PIK3CA", "GATA3"
)

TARGET_GENES <- c(
    "ERBB2",   # HER2 — trastuzumab, pertuzumab, T-DM1, trastuzumab deruxtecan
    "ESR1",    # ER — tamoxifen, aromatase inhibitors, fulvestrant, SERDs
    "PIK3CA",  # PI3K — alpelisib
    "CDK4",    # CDK4/6 inhibitors — palbociclib, ribociclib, abemaciclib
    "CDK6",    # same class as above
    "PARP1",   # PARP inhibitors (olaparib, talazoparib) in BRCA1/2-mutant tumors
    "AKT1",    # AKT inhibitors — capivasertib
    "MTOR",    # mTOR inhibitors — everolimus
    "TOP1",    # topoisomerase I — target of deruxtecan payload in ADCs
    "TACSTD2" # Trop-2 — target of sacituzumab govitecan
)

CONDITION = "cond_a"
CONDITION_A = "pos"
CONDITION_B = "neg"

BATCHES = c("all")

if (!length(BATCHES) %in% c(1, 2)) {
  stop("BATCHES must have exactly 1 or 2 elements; got ", length(BATCHES))
}

# Batch is removed upstream by ComBat-seq rather than modelled here, so the
# design is the condition alone in both the one- and two-batch cases.
#
# This is only valid because ComBat-seq returns count-like non-negative integers
# and was run with `group = CONDITION`, which preserves the biological signal of
# interest. If the adjusted matrix is ever regenerated without `group`, condition
# effects that align with batch will have been removed and this design will be
# testing a matrix with some of the signal already deleted.
DESIGN_FORMULA <- reformulate(CONDITION)

# Quarto renders with the working directory set to this file's folder, but
# interactive console runs start at the project root, so resolve `data/` either way.
DATA_DIR <- if (dir.exists("data")) "data" else "../data"

COUNT_TABLE_FILE <- file.path(DATA_DIR, "count_table_demo.csv")

# This document reads the count matrix named above as-is; the flag records
# whether that file is a ComBat-seq adjusted matrix (produced upstream and named
# adjusted_counts_<condition>.csv) rather than the raw counts.
USE_ADJUSTED_COUNTS <- grepl("^adjusted_counts", basename(COUNT_TABLE_FILE))

if (length(BATCHES) == 2 && !USE_ADJUSTED_COUNTS) {
  warning("Two batches requested but COUNT_TABLE_FILE is the raw matrix; ",
          "batch is not in DESIGN_FORMULA, so batch effects are unmodelled.",
          call. = FALSE)
}

METADATA_FILE <- file.path(DATA_DIR, "metadata_demo.csv")

# Cached Ensembl gene annotation; see src/fetch_annotation.R for how it is built.
# Used only as the source of `gene_biotype`.
ANNOTATION_FILE <- file.path(DATA_DIR, "annotation.csv")

# RSEM output carrying the gene_id and gene_name from the GTF the FASTQs were
# quantified against; the authoritative ID -> symbol map for this count matrix.
RSEM_COUNTS_FILE <- file.path(DATA_DIR, "rsem.merged.gene_counts.tsv")

# GTF symbols joined to Ensembl biotypes; see src/build_annotation_updated.R.
ANNOTATION_UPDATED_FILE <- file.path(DATA_DIR, "annotation_updated.csv")

# --- Output locations -------------------------------------------------------
# Resolved the same way as DATA_DIR so the paths work both when Quarto renders
# from analysis/ and when chunks are run interactively from the project root.
OUT_DIR <- if (dir.exists("data")) "outputs" else "../outputs"
FIG_DIR <- file.path(OUT_DIR, "figures")
TAB_DIR <- file.path(OUT_DIR, "tables")

for (d in c(OUT_DIR, FIG_DIR, TAB_DIR)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}

# Figure indices are passed in explicitly rather than taken from a running
# counter, so a name is stable even when a single chunk is re-run on its own.
save_figure <- function(plot, index, name, width = 8, height = 6, dpi = 300) {
  if (is.null(plot)) {
    message("no figure to save for ", name)
    return(invisible(NULL))
  }
  path <- file.path(FIG_DIR, sprintf("%02d_%s.png", index, name))

  # pheatmap returns a list; its $gtable is the drawable object.
  obj <- if (inherits(plot, "pheatmap")) plot$gtable else plot

  if (inherits(obj, c("ggplot", "patchwork"))) {
    ggplot2::ggsave(path, obj, width = width, height = height, dpi = dpi)
  } else {
    # grid objects (pheatmap gtable, fgsea::plotGseaTable) need a device
    png(path, width = width, height = height, units = "in", res = dpi)
    on.exit(dev.off(), add = TRUE)
    grid::grid.newpage()
    grid::grid.draw(obj)
  }
  invisible(path)
}

save_table <- function(x, name) {
  path <- file.path(TAB_DIR, name)
  readr::write_csv(x, path)
  invisible(path)
}

```

# Metadata

```{r}

metadata_all <- read_csv(METADATA_FILE)
metadata_all <- column_to_rownames(metadata_all, var = "slid")
head(metadata_all)
metadata <- metadata_all[,c("batch", CONDITION)]
tail(metadata)

```

```{r}
if (!file.exists(COUNT_TABLE_FILE)) {
  stop("Count table not found: ", COUNT_TABLE_FILE,
       if (USE_ADJUSTED_COUNTS)
         "\nTwo batches requested, so a ComBat-seq matrix for CONDITION = '"
       else "",
       if (USE_ADJUSTED_COUNTS) paste0(CONDITION, "' must be generated first.") else "")
}

# The adjusted matrix is condition-specific and generated outside this document,
# so warn if it predates the metadata it is supposed to match.
# if (USE_ADJUSTED_COUNTS &&
#     file.exists(METADATA_FILE) &&
#     file.mtime(COUNT_TABLE_FILE) < file.mtime(METADATA_FILE)) {
#   warning("ComBat-seq matrix is older than the metadata; it may be stale. ",
#           "Regenerate it with group = '", CONDITION, "' if samples changed.",
#           call. = FALSE)
# }

# message("Counts: ", basename(COUNT_TABLE_FILE),
#         if (USE_ADJUSTED_COUNTS) "  [ComBat-seq adjusted]" else "  [raw]")

count_table_raw <- read.csv(COUNT_TABLE_FILE, header = TRUE, sep = ",", check.names = FALSE)

colnames(count_table_raw)[1] = "gene_id"

# Drop genes with zero counts in every sample; they carry no information and
# only inflate the multiple-testing burden downstream.
keep_nonzero <- rowSums(count_table_raw[, -1, drop = FALSE], na.rm = TRUE) > 0
message(sum(!keep_nonzero), " of ", nrow(count_table_raw),
        " genes had zero counts in all samples and were removed.")
count_table_raw <- count_table_raw[keep_nonzero, , drop = FALSE]

# Row subsetting leaves behind the original integer row names, which
# column_to_rownames() below rejects as pre-existing row names.
rownames(count_table_raw) <- NULL

# Row subsetting leaves behind the original integer row names, which
# column_to_rownames() below rejects as pre-existing row names.
rownames(count_table_raw) <- NULL

count_table <- column_to_rownames(count_table_raw, var = "gene_id")

```

# avatar and cicpt with CONDITION status

```{r}
metadata <- metadata %>% dplyr::filter(!is.na(.data[[CONDITION]]), batch %in% BATCHES)

# Subsetting by a name that is absent yields an all-NA column rather than an
# error, so check membership before indexing.
missing_samples <- setdiff(rownames(metadata), colnames(count_table))
if (length(missing_samples) > 0) {
  stop(length(missing_samples), " samples in the metadata are absent from ",
       basename(COUNT_TABLE_FILE), ": ",
       paste(head(missing_samples, 10), collapse = ", "))
}

count_table <- count_table[, rownames(metadata)]
```



# table for CONDITION and batch
```{r}
table(metadata$batch, metadata[[CONDITION]])
```


# Checking duplicated rows after clean gene id. They are identical. Delete duplicates in the next chunk.
```{r}
gene_ids <- rownames(count_table)
clean_gene_ids <- sub("\\..*$", "", gene_ids)
dup_ids <- unique(clean_gene_ids[duplicated(clean_gene_ids)])
length(dup_ids)

dup_rows <- count_table[clean_gene_ids %in% dup_ids, ]
dup_rows_ordered <- dup_rows[order(sub("\\..*$", "", rownames(dup_rows))), ]
head(dup_rows_ordered, 12)
```

# Drop the duplicated rows

The first row of each duplicated base ID is kept and the rest dropped, and
versions are stripped here, once, rather than relying on deduplication later.

Which row survives is only safe to leave to file order because the rows are first
asserted to be identical: every duplicate row is compared to the first row of its
own group, so if a future count matrix ever assigns different values to the copies
this fails loudly instead of silently discarding signal. In this dataset the
duplicates are GENCODE pseudoautosomal genes, emitted once on X and again as a
`_PAR_Y` copy, but nothing here depends on that.

```{r}
gene_ids <- rownames(count_table)
clean_gene_ids <- sub("\\..*$", "", gene_ids)

dup <- clean_gene_ids %in% clean_gene_ids[duplicated(clean_gene_ids)]
dup_mat <- as.matrix(count_table[dup, , drop = FALSE])

# Row index of the first member of each duplicate group, recycled to compare
# every member against it. Works for groups of any size.
first_of_group <- match(clean_gene_ids[dup], clean_gene_ids[dup])
stopifnot(all(dup_mat == dup_mat[first_of_group, , drop = FALSE]))

keep <- !duplicated(clean_gene_ids)
count_table <- count_table[keep, ]
rownames(count_table) <- clean_gene_ids[keep]

stopifnot(!any(duplicated(rownames(count_table))))

dim(count_table)
```



# Gene id - Gene symbol - gene types

Row names are already stripped to base Ensembl IDs by the chunk above, so they
join directly to the annotation.

Symbols come from the GTF the FASTQs were quantified against, read out of the
RSEM merged counts file. That covers every gene in the count matrix, including
~1,290 IDs that Ensembl has since retired and which a current BioMart query
cannot resolve. Biotypes are not in the GTF output, so they are joined in from
the cached Ensembl table — see [src/build_annotation_updated.R].

`useEnsembl()` no longer works for that biotype table: Ensembl retired the
classic BioMart path on `www.ensembl.org` and decommissioned the `useast`/`asia`
mirrors, which are the only hosts biomaRt 2.62 will try. It is therefore fetched
from a dated BioMart archive and cached — see [src/fetch_annotation.R].

Note that symbols are not unique (~1,500 `gene_name` values are duplicated in the
GTF), so `ensembl_gene_id` stays the key for every join and for rownames.

```{r, message = FALSE}
source(file.path(SRC_DIR, "fetch_annotation.R"))
source(file.path(SRC_DIR, "build_annotation_updated.R"))

# Biotype source only. Downloads on first run, reads the CSV afterwards.
annotation <- load_annotation(ANNOTATION_FILE, refresh = FALSE)

# Rebuild the merged table whenever either input is newer than the cache.
annotation_stale <- !file.exists(ANNOTATION_UPDATED_FILE) ||
  file.mtime(ANNOTATION_UPDATED_FILE) <
    max(file.mtime(c(RSEM_COUNTS_FILE, ANNOTATION_FILE)))

annotation_updated <- load_annotation_updated(
  ANNOTATION_UPDATED_FILE,
  rsem_file = RSEM_COUNTS_FILE,
  biotype_file = ANNOTATION_FILE,
  refresh = annotation_stale
)

stopifnot(
  !any(duplicated(annotation_updated$ensembl_gene_id)),
  all(rownames(count_table) %in% annotation_updated$ensembl_gene_id)
)

# Retired IDs keep their GTF symbol but have no current biotype, so the filter
# below drops them; say so rather than letting them vanish silently.
n_unannotated <- sum(is.na(annotation_updated$gene_biotype))
if (n_unannotated > 0) {
  message(n_unannotated, " of ", nrow(annotation_updated),
          " genes have a GTF symbol but no current Ensembl biotype (retired IDs); ",
          "they are excluded by the biotype filter below.")
}

head(annotation_updated, 5)
```


# OLD - Gene id - Gene symbol - gene types
```{r, message = FALSE}
# gene_ids <- rownames(count_table)
# clean_gene_ids <- sub("\\..*$", "", gene_ids)
# 
# count_table <- count_table[!duplicated(clean_gene_ids), ]
# rownames(count_table) <- clean_gene_ids[!duplicated(clean_gene_ids)]
# 
# ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
#                                                   # mouse - "mmusculus_gene_ensembl", human - "hsapiens_gene_ensembl"
# 
# # human 'hgnc_symbol', mouse 'mgi_symbol'
# # retrieve gene symbols, gene types
# annotation <- getBM(
#   attributes = c('ensembl_gene_id', 'hgnc_symbol', 'gene_biotype'),
#   filters = 'ensembl_gene_id',
#   values = clean_gene_ids,
#   mart = ensembl
# )
# 
# empty_index = which(annotation$hgnc_symbol=="")
# annotation$hgnc_symbol[empty_index]=annotation$ensembl_gene_id[empty_index]
# 
# write_csv(annotation, "annotation.csv")

# annotation <- read.csv("../annotation.csv", header = TRUE, stringsAsFactors = FALSE)

# n_dup <- sum(duplicated(annotation$ensembl_gene_id))
# if (n_dup > 0) message("Dropping ", n_dup, " duplicate ensembl_gene_id rows from annotation")
#
# annotation <- annotation |> distinct(ensembl_gene_id, .keep_all = TRUE)
# stopifnot(!any(duplicated(annotation$ensembl_gene_id)))
#
# head(annotation,5)
```

# types of genes in the data

Counted over the genes actually present in the count matrix, with retired IDs
shown as `NA`.

```{r}
annotation_updated %>%
  filter(ensembl_gene_id %in% rownames(count_table)) %>%
  count(gene_biotype, sort = TRUE) %>%
  head(10)
```

# Protein coding genes only

`gene_selected` is intersected with the count table rownames. Indexing a data
frame by absent rownames returns silent all-`NA` rows rather than an error, and
the annotation covers genes that are not in this matrix.

```{r}
gene_type_pick = GENE_TYPE_PICK
print(gene_type_pick)

gene_selected <- intersect(
  rownames(count_table),
  annotation_updated$ensembl_gene_id[annotation_updated$gene_biotype %in% gene_type_pick]
)
length(gene_selected)
```



# Trim the count table for selected genes categories e.g. protein coding only
```{r}
data <- count_table[gene_selected,]
stopifnot(!anyNA(data), nrow(data) == length(gene_selected))
dim(data)
head(data[1:5, 1:10])
```


# the total reads per sample
```{r}
n <- ncol(data)
colSums(data)[c(1:10, (n - 4):n)]
```

# Checking sample id in data & metadata
```{r}
# metadata <- metadata[colnames(data), ]

# DESeq2 pairs count columns to metadata rows by position, not by name, so a
# set or order mismatch silently assigns the wrong condition to samples instead
# of erroring. Each check is therefore fatal.
extra_meta <- setdiff(rownames(metadata), colnames(data))
if (length(extra_meta) > 0) {
  stop(length(extra_meta), " sample(s) in metadata are absent from the count matrix: ",
       paste(head(extra_meta, 10), collapse = ", "), call. = FALSE)
}

extra_data <- setdiff(colnames(data), rownames(metadata))
if (length(extra_data) > 0) {
  stop(length(extra_data), " sample(s) in the count matrix are absent from metadata: ",
       paste(head(extra_data, 10), collapse = ", "), call. = FALSE)
}

# Reaching here means the two sets match, so only the ordering can differ.
if (!identical(colnames(data), rownames(metadata))) {
  stop("Count matrix columns and metadata rows hold the same samples in ",
       "different orders; uncomment the reorder above or fix it upstream.",
       call. = FALSE)
}

ncol(data)
```

# Setup DESeq2
```{r, message = FALSE}
metadata[[CONDITION]] <- factor(metadata[[CONDITION]], levels = c(CONDITION_A, CONDITION_B))
# metadata[[CONDITION]] <- relevel(metadata[[CONDITION]], ref = "FALSE")  # be explicit
metadata$batch <- factor(metadata$batch)       

# Both condition levels must be present, or the contrast is not estimable.
mm <- model.matrix(DESIGN_FORMULA, data = metadata)
stopifnot(qr(mm)$rank == ncol(mm))

# ComBat-seq should return non-negative integers. Anything else means the matrix
# came from a method whose output DESeq2's NB model cannot legitimately consume
# (e.g. limma::removeBatchEffect on log-scale data).
count_mat <- as.matrix(data)
stopifnot(is.numeric(count_mat))
if (any(count_mat < 0, na.rm = TRUE)) {
  stop("Negative values in the count matrix: not usable as counts.")
}

dds <- DESeqDataSetFromMatrix(countData = round(as.matrix(data)),
                              colData = metadata,
                              design = DESIGN_FORMULA)

dds <- DESeq(dds)
```


# variance stabilizing transformation
```{r}
vsd <- vst(dds, blind = TRUE)
```

# Before (vst) and After (vsd)
```{r}
library(vsn)   # provides meanSdPlot

# BEFORE: shifted log2 of normalized counts
ntd <- normTransform(dds)          # log2(x + 1) on normalized counts
msd_before <- meanSdPlot(assay(ntd))   # SD tends to rise at low/high mean
save_figure(msd_before$gg, 0, "meansd_before_vst", width = 6, height = 5)

# AFTER: variance stabilizing transform
vsd <- vst(dds, blind = TRUE)
msd_after <- meanSdPlot(assay(vsd))
save_figure(msd_after$gg, 1, "meansd_after_vst", width = 6, height = 5)
```

# Distance Plot

The distance plot calculates the Euclidean distance between the
expression values for each individual sample. This is a way to identify
which samples are most closely related in terms of their gene expression
patterns.


```{r}
save_figure(plotDists(vsd, label_col = CONDITION, fontsize = 6),
            2, "sample_distance_heatmap", width = 10, height = 9)
```


# Variable Genes Heatmap.

### Working version

```{r}

save_figure(variable_gene_heatmap(vsd, num_genes = 50, annotation = annotation),
            3, "variable_gene_heatmap", width = 10, height = 9)
```


# PCA Plot

```{r}
# plot_PCA <- function(vsd.obj, group_var) {
#   pcaData <- plotPCA(vsd.obj, intgroup = group_var, returnData = TRUE)
#   percentVar <- round(100 * attr(pcaData, "percentVar"))
#   ggplot(pcaData, aes(PC1, PC2, color = .data[[group_var]])) +
#     geom_point(size = 3) +
#     labs(x = paste0("PC1: ", percentVar[1], "% variance"),
#          y = paste0("PC2: ", percentVar[2], "% variance"),
#          title = paste0("PCA Plot colored by ", group_var)) +
#     ggrepel::geom_text_repel(aes(label = name), color = "black", size = 2)
# }

save_figure(plot_PCA(vsd, CONDITION), 4, "pca_condition", width = 7, height = 6)
plot_PCA(vsd, CONDITION)
```

When two batches are analysed the counts have already been through ComBat-seq,
so this plot doubles as a check that the correction worked: batches should be
intermingled rather than separating along PC1 or PC2. Residual separation here
means batch is still driving the data despite adjustment, and the `~ condition`
design would then be unable to account for it.

```{r}
save_figure(plot_PCA(vsd, "batch"), 5, "pca_batch", width = 7, height = 6)
plot_PCA(vsd, "batch")
```



```{r}
# res <- results(dds)

all_genes_results <- results(dds,
                             contrast = c(CONDITION, # name of the factor
                                          CONDITION_A, # numerator level for fold change
                                          CONDITION_B),
                             # independent filtering is optimised for this alpha,
                             # so it must match the threshold used downstream
                             alpha = PADJ_THRES)
all_diff_genes <- all_genes_results %>% as.data.frame()

all_diff_genes_df <- all_genes_results %>% as.data.frame() %>%
  rownames_to_column("ensembl_id") %>%
  left_join(annotation %>% dplyr::select(ensembl_gene_id, hgnc_symbol),
            by = c("ensembl_id" = "ensembl_gene_id")) %>%
  arrange(padj, desc(log2FoldChange))

save_table(all_diff_genes_df, "all_diff_genes_df.csv")
```

------------------------------------------------------------------------

## If things are good,

- **1. High Data Quality & Balanced Expression Baseline** The vast
  majority of non-differentially expressed genes (grey points) form a
  symmetric, tight cloud directly centered along
  $\text{log}_2\text{ fold change} = 0$. This confirms that sample
  normalization was successful and that there is no global baseline
  bias.
- **2. Predominantly Up-Regulated Transcriptional Response** While
  differentially expressed genes (red points) are detected in both
  directions, there is a clear bias toward **up-regulation** in the
  treatment condition. Significant genes above the dashed line outnumber
  those below it and reach higher absolute magnitude (peaks up to
  $\sim 10$ $\text{log}_2\text{ FC}$).
- **3. Robust Statistical Control Over Low-Count Noise** At low
  expression levels (left side, near $1\text{e}+00$), the plot displays
  expected Poisson variance ("fan" shape), but virtually all of these
  high-variance points are correctly identified as **non-significant**
  (grey). Statistical power increases with read coverage, allowing true
  differentially expressed genes to be called reliably across
  moderate-to-high mean expression levels ($10^1$ to $10^5$).

```{r}
# plotMA(all_genes_results, alpha = PADJ_THRES, ylim = c(-5, 5))

all_diff_genes_df <- all_diff_genes_df %>%
  mutate(significant = !is.na(padj) & padj < PADJ_THRES)

ma_plot <- ggplot(all_diff_genes_df, aes(x = baseMean, y = log2FoldChange, color = significant)) +
  geom_point(alpha = 0.5, size = 0.8) +
  scale_x_log10() +
  scale_color_manual(values = c("FALSE" = "grey60", "TRUE" = "red")) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Mean of normalized counts", y = "log2 fold change",
       title = "MA plot: etf_12 pos vs neg") +
  theme_minimal()

save_figure(ma_plot, 6, "ma_plot", width = 7, height = 5)
ma_plot

```
***google: In MA plot of mean normalized count vs log2 fold change, what is the significance?***


# Differentially expressed genes

padj \< `{r} PADJ_THRES`

```{r}
diff_genes <- all_genes_results %>% 
              as.data.frame() %>%
              # rownames_to_column("genes") %>%
              filter(padj < PADJ_THRES) %>%
              arrange(padj, desc(log2FoldChange))
diff_genes <- rownames_to_column(diff_genes, var = "ensembl_id")
diff_genes <- left_join(diff_genes, annotation, by = c("ensembl_id" = "ensembl_gene_id"))
save_table(diff_genes, "diff_genes_significant.csv")
head(diff_genes)

```

# Checking TARGET genes
```{r}
mask <- all_diff_genes_df$hgnc_symbol %in% c(TARGET_GENES) # KIT = CD117
target_gene_list <- all_diff_genes_df[mask,c("hgnc_symbol", "ensembl_id", "padj", "pvalue")] %>% arrange(padj, pvalue)

target_gene_list

# --- Arrange in pages of 30 (6 cols x 5 rows) ---
n_genes    <- nrow(target_gene_list)
page_size  <- 30
n_pages    <- ceiling(n_genes / page_size)

# for (page in seq_len(n_pages)) {
#   idx   <- ((page - 1) * page_size + 1):min(page * page_size, n_genes)
#   # plots <- lapply(idx, boxplot_gene_counts(sig_gene_list, dds, group = CONDITION))
#   plots <- lapply(idx, boxplot_gene_counts,
#                 genes = target_gene_list, dds = dds, group = CONDITION)
# 
#   p <- wrap_plots(plots, ncol = 3, nrow = 3)
# 
#   ggsave(
#     filename = sprintf("target_gene_counts.png"),
#     plot     = p,
#     width    = 6,   # inches; ~2.3" per panel across 6 cols
#     height   = 6,   # inches; ~2.4" per panel across 5 rows
#     device   = "png"
#   )
# }

for (page in seq_len(n_pages)) {
  idx <- ((page - 1) * page_size + 1):min(page * page_size, n_genes)

  plots <- lapply(idx, function(i) {
    gene_row <- target_gene_list[i, ]
    subtitle <- sprintf("padj = %.3g | p = %.3g", gene_row$padj, gene_row$pvalue)

    boxplot_gene_counts(i, genes = target_gene_list, dds = dds, group = CONDITION) +
      labs(subtitle = subtitle)
  })

  # A fixed nrow x ncol errors out when the gene list does not fit the grid,
  # so set the columns and let the row count follow the panel count.
  n_col <- 3
  n_row <- ceiling(length(plots) / n_col)
  p <- wrap_plots(plots, ncol = n_col)

  # One file per page; with a single page the suffix is dropped.
  fig_name <- if (n_pages > 1) sprintf("target_gene_counts_p%d", page) else "target_gene_counts"
  save_figure(p, 7, fig_name, width = 2.7 * n_col, height = 2.4 * n_row)
}

p
print("Check the saved PNG file.")
```


# Checking Signature genes + C200 + CD117 (KIT)
```{r}
mask <- all_diff_genes_df$hgnc_symbol %in% c(SIGNATURE_GENES, c("CD200", "KIT")) # KIT = CD117
sig_gene_list <- all_diff_genes_df[mask,c("hgnc_symbol", "ensembl_id", "padj", "pvalue")] %>% arrange(padj, pvalue)

sig_gene_list


# --- Arrange in pages of 30 (6 cols x 5 rows) ---
n_genes    <- nrow(sig_gene_list)
page_size  <- 30
n_pages    <- ceiling(n_genes / page_size)

for (page in seq_len(n_pages)) {
  idx   <- ((page - 1) * page_size + 1):min(page * page_size, n_genes)
  # plots <- lapply(idx, boxplot_gene_counts(sig_gene_list, dds, group = CONDITION))
  plots <- lapply(idx, boxplot_gene_counts,
                genes = sig_gene_list, dds = dds, group = CONDITION)

  n_col <- 4
  n_row <- ceiling(length(plots) / n_col)
  p <- wrap_plots(plots, ncol = n_col)

  fig_name <- if (n_pages > 1) sprintf("sig_gene_counts_p%d", page) else "sig_gene_counts"
  save_figure(p, 8, fig_name, width = 2.7 * n_col, height = 2.4 * n_row)
}
p
print("Check the saved PNG file.")
```

# PlotCounts - top eight genes

We are often interested in plotting the normalized counts for certain
genes of interest, in order to get a sense of the spread of the data and
how genes are expressed across each of the groups. This is especially of
interest for genes that show up as differentially expressed in our
DESeq2 results. A way to plot the normalized counts for each sample is
by using the built-in DESeq2::plotCounts() function. However, this
function leaves a lot to be desired. As an example, I plot the gene
IGFBP1 (a hypoxia-inducible gene) using its Ensembl ID below.

```{r}

# importing plot_gene_counts <- function(ii, genes, dds, group)
# source("../R_scripts/RNAseq_helpers.R")

# plots <- lapply(1:8, plot_gene_counts)
plots <- lapply(1:8, plot_gene_counts,
                genes = diff_genes, dds = dds, group = CONDITION)

save_figure((plots[[1]] + plots[[2]]) / (plots[[3]] + plots[[4]]),
            9, "top_gene_counts_1_4", width = 9, height = 7)
save_figure((plots[[5]] + plots[[6]]) / (plots[[7]] + plots[[8]]),
            10, "top_gene_counts_5_8", width = 9, height = 7)

(plots[[1]] + plots[[2]]) / (plots[[3]] + plots[[4]])

(plots[[5]] + plots[[6]]) / (plots[[7]] + plots[[8]])
```


```{r}
all_diff_genes <- na.omit(all_diff_genes)

all_diff_genes$significance = as.factor(all_diff_genes$padj < PADJ_THRES)

all_diff_genes <- rownames_to_column(all_diff_genes, var = 'ensembl_id')

all_diff_genes <- left_join(all_diff_genes, annotation, by = c("ensembl_id" = "ensembl_gene_id"))
save_table(all_diff_genes, "all_diff_genes.csv")

top_genes <- all_diff_genes %>% arrange(padj) %>% head(NLABEL)
```

# Volcano
```{r}
volcano_plot <- ggplot(all_diff_genes, aes(log2FoldChange, -log10(padj))) +
  geom_point(aes(col=significance), size=0.5) + 
  scale_color_manual(values=c("black", "red")) +
  ggrepel::geom_text_repel(data=top_genes, aes(label=hgnc_symbol), size = 3, max.overlaps=13) +
  labs(x = "Log2FoldChange", y = "-(Log10 adjusted p-value)") +
  geom_vline(xintercept = 0, linetype = "dotted") + 
  geom_hline(yintercept = -log10(PADJ_THRES), linetype = "dotted") +
  theme_minimal()

save_figure(volcano_plot, 11, "volcano", width = 8, height = 6)
volcano_plot
```
# Gene Ontology (GO)
| Abbreviation | Full Name | Describes |
| --- | --- | --- |
| BP | Biological Process | What the gene does in a biological context — e.g., "cell cycle regulation", "apoptosis", "immune response" |
| CC | Cellular Component | Where the gene product is located — e.g., "nucleus", "mitochondria", "plasma membrane" |
| MF | Molecular Function | The biochemical activity of the gene product — e.g., "kinase activity", "DNA binding", "receptor ligand" |


# Enrichment background

The default `enrichGO()` background is every annotated gene in the OrgDb, which
is much larger than the set actually tested here (protein-coding genes expressed
in this cohort). That inflates terms enriched for expressed genes generally. The
universe below is the tested gene set, mapped through the same
`SYMBOL -> ENTREZID` route as the DE list so numerator and denominator are
comparable.

```{r, message = FALSE}
universe_symbols <- all_diff_genes_df %>%
  dplyr::filter(!is.na(pvalue)) %>%
  dplyr::pull(hgnc_symbol) %>%
  unique()

GO_UNIVERSE <- bitr(
  geneID   = universe_symbols,
  fromType = "SYMBOL",
  toType   = "ENTREZID",
  OrgDb    = org.Hs.eg.db
)$ENTREZID

GO_UNIVERSE <- unique(GO_UNIVERSE)
length(GO_UNIVERSE)
```

# BP: Biological Process enrichGO. Expressed genes as background
```{r, message = FALSE}
onto = "BP"

my_gene_list <- diff_genes$hgnc_symbol
converted_genes <- bitr(
  geneID = my_gene_list,
  fromType = "SYMBOL", # Change this to "ENSEMBL" if that's your starting format
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

# every DE gene must be inside the background it is tested against
stopifnot(all(converted_genes$ENTREZID %in% GO_UNIVERSE))

go_results <- enrichGO(
  gene         = converted_genes$ENTREZID,
  universe     = GO_UNIVERSE,
  OrgDb        = org.Hs.eg.db,
  keyType      = "ENTREZID", # Important to specify the input ID type
  ont          = onto,
  pvalueCutoff = PVALUECUTOFF,
  pAdjustMethod = "BH", # Benjamini-Hochberg correction for multiple testing
  readable     = TRUE # Converts Entrez IDs back to gene symbols for readability
)

if ( !is.null(go_results)  ){
  
  if (dim(go_results)[1] > 0) {
    go_bar <- barplot(go_results,
                      showCategory = min(dim(go_results)[1], ENRICH_BARCOUNT),
                      font.size = 9)
    save_figure(go_bar, 12, paste0("enrichgo_", onto), width = 8, height = 7)
    print(go_bar)
    go_df <- as.data.frame(go_results)
    save_table(go_df, paste0("go_df_", onto, ".csv"))
  }
  
  if ( dim(go_results)[1] == 0 ) {
    print('No enrichGO result')
  }

}


```

```{r}
# The helper writes its own CSV; point it at the tables folder and take over
# plotting so the barplot can be saved with an indexed filename.
go_enrichment_with_figure <- function(onto, index) {
  go <- run_go_enrichment(diff_genes, onto, universe = GO_UNIVERSE,
                          out_dir = TAB_DIR, plot = FALSE)
  n_terms <- if (is.null(go)) 0L else nrow(as.data.frame(go))
  if (n_terms == 0) return(invisible(go))

  bar <- barplot(go, showCategory = min(n_terms, ENRICH_BARCOUNT), font.size = 9)
  save_figure(bar, index, paste0("enrichgo_", onto), width = 8, height = 7)
  print(bar)
  invisible(go)
}

go_enrichment_with_figure("BP", 12)
```

## CC: Cellular Component enrichGO. Expressed genes as background
```{r}
go_enrichment_with_figure("CC", 13)
```


## MF: Molecular Function enrichGO. Expressed genes as background
```{r}
go_enrichment_with_figure("MF", 14)
```


# MsigDB categories

| Category | Name | Description |
|------------------------|------------------------|------------------------|
| H | Hallmark | Well-defined biological states/processes, derived by summarizing multiple related gene sets |
| C1 | Positional | Gene sets by chromosome/cytogenetic band |
| C2 | Curated | From pathway databases (e.g., KEGG, Reactome, BioCarta) and published studies; includes CGP (chemical/genetic perturbations) and CP (canonical pathways) subsets |
| C3 | Regulatory targets | Motif-based sets — transcription factor targets (TFT) and microRNA targets (MIR) |
| C4 | Computational | Cancer-oriented gene sets defined by mining large expression datasets (CGN, CM) |
| C5 | Ontology | Gene Ontology (GO: BP, CC, MF) and Human Phenotype Ontology (HPO) |
| C6 | Oncogenic signatures | Gene sets from perturbations of cancer-related pathways |
| C7 | Immunologic signatures | From immune-related studies (ImmuneSigDB) and immune response gene sets (VAX) |
| C8 | Cell type signatures | Cluster/cell-type markers from single-cell studies |

```{r}
library(msigdbr)
library(ggplot2)
```

```{r}
res_ranked <- all_genes_results %>% as.data.frame() %>%
  rownames_to_column("ensembl_id") %>%
  mutate(rank_metric = -log10(pvalue) * sign(log2FoldChange)) %>%
  drop_na(rank_metric) %>%
  # fgsea sorts `stats` itself and orders tied values arbitrarily, so a tied
  # ranking makes the GSEA result irreproducible. The ties are almost entirely
  # near-zero-expression genes sharing an identical Wald p-value. Order the table
  # deterministically -- metric, then expression, then gene ID -- and subtract a
  # rank-proportional decrement so the metric is strictly decreasing.
  #
  # The decrement is applied to every gene rather than only to tied ones because
  # some distinct metric values differ by less than a double's precision at that
  # magnitude (min gap ~4e-19), so a within-tie epsilon is silently absorbed.
  # Total distortion is under 1e-5, negligible against a metric range of ~17 and
  # the |stat|^gseaParam weighting fgsea applies.
  arrange(desc(rank_metric), desc(baseMean), ensembl_id) %>%
  mutate(rank_metric = rank_metric - row_number() * RANK_TIE_DELTA)

stopifnot(all(diff(res_ranked$rank_metric) < 0))
```

```{r}
# Create the named vector required by fgsea
gene_ranks <- res_ranked$rank_metric
names(gene_ranks) <- res_ranked$ensembl_id

names(gene_ranks) <- gsub("\\..*$", "", names(gene_ranks))
```

# MSigDB human: H, C1 \~ C8
## MsigDB - H
```{r}
category = "H"
msigdb_X <- msigdb_function(category,gene_ranks)

save_table(msigdb_X[[1]], paste0("msigdb_results_", category, "_test.csv"))
save_figure(msigdb_X$table_plot, 15, paste0("msigdb_", category, "_table"),
            width = 11, height = 6)
save_figure(msigdb_X$enrichment_plots, 16, paste0("msigdb_", category, "_enrichment"),
            width = 10, height = 8)

# 10 upregulated, 10 downregulated
msigdb_X[[2]]

# Enrichment plots for 2 top-up and 2 top-down pathways
msigdb_X[[3]]
```

## MsigDB - C4

```{r}
category = "C4"
msigdb_X <- msigdb_function(category,gene_ranks)

save_table(msigdb_X[[1]], paste0("msigdb_results_", category, "_test.csv"))
save_figure(msigdb_X$table_plot, 17, paste0("msigdb_", category, "_table"),
            width = 11, height = 6)
save_figure(msigdb_X$enrichment_plots, 18, paste0("msigdb_", category, "_enrichment"),
            width = 10, height = 8)

# 10 upregulated, 10 downregulated
msigdb_X[[2]]

# Enrichment plots for 2 top-up and 2 top-down pathways
msigdb_X[[3]]
```


## MsigDB - C5

```{r}
category = "C5"
msigdb_X <- msigdb_function(category,gene_ranks)

save_table(msigdb_X[[1]], paste0("msigdb_results_", category, "_test.csv"))
save_figure(msigdb_X$table_plot, 19, paste0("msigdb_", category, "_table"),
            width = 11, height = 6)
save_figure(msigdb_X$enrichment_plots, 20, paste0("msigdb_", category, "_enrichment"),
            width = 10, height = 8)

# 10 upregulated, 10 downregulated
msigdb_X[[2]]

# Enrichment plots for 2 top-up and 2 top-down pathways
msigdb_X[[3]]
```


## MsigDB - C6

```{r}
category = "C6"
msigdb_X <- msigdb_function(category,gene_ranks)

save_table(msigdb_X[[1]], paste0("msigdb_results_", category, "_test.csv"))
save_figure(msigdb_X$table_plot, 21, paste0("msigdb_", category, "_table"),
            width = 11, height = 6)
save_figure(msigdb_X$enrichment_plots, 22, paste0("msigdb_", category, "_enrichment"),
            width = 10, height = 8)

# 10 upregulated, 10 downregulated
msigdb_X[[2]]

# Enrichment plots for 2 top-up and 2 top-down pathways
msigdb_X[[3]]
```


## MsigDB - C7

```{r}
category = "C7"
msigdb_X <- msigdb_function(category,gene_ranks)

save_table(msigdb_X[[1]], paste0("msigdb_results_", category, "_test.csv"))
save_figure(msigdb_X$table_plot, 23, paste0("msigdb_", category, "_table"),
            width = 11, height = 6)
save_figure(msigdb_X$enrichment_plots, 24, paste0("msigdb_", category, "_enrichment"),
            width = 10, height = 8)

# 10 upregulated, 10 downregulated
msigdb_X[[2]]

# Enrichment plots for 2 top-up and 2 top-down pathways
msigdb_X[[3]]
```

## MsigDB - C8

```{r}
category = "C8"
msigdb_X <- msigdb_function(category,gene_ranks)

save_table(msigdb_X[[1]], paste0("msigdb_results_", category, "_test.csv"))
save_figure(msigdb_X$table_plot, 25, paste0("msigdb_", category, "_table"),
            width = 11, height = 6)
save_figure(msigdb_X$enrichment_plots, 26, paste0("msigdb_", category, "_enrichment"),
            width = 10, height = 8)

# 10 upregulated, 10 downregulated
msigdb_X[[2]]

# Enrichment plots for 2 top-up and 2 top-down pathways
msigdb_X[[3]]
```
```{r}
new_file_name = paste0("RNAseq_DESeq2_with_data_", paste(BATCHES, collapse = "_"), "_groupby_",CONDITION,".html")
if (file.exists("RNAseq_DESeq2_template.html")) {
  file.rename("RNAseq_DESeq2_template.html", new_file_name)
}
```


# Summary Report
```{r}
# ============================================================================
# Summary Report
# ============================================================================
output_dir <- OUT_DIR

# --- Cohort counts ---
n_samples <- nrow(metadata)
n_a <- sum(metadata[[CONDITION]] == CONDITION_A, na.rm = TRUE)
n_b <- sum(metadata[[CONDITION]] == CONDITION_B, na.rm = TRUE)

# --- Signature / target gene coverage ---
sig_present <- intersect(SIGNATURE_GENES, all_diff_genes_df$hgnc_symbol)
sig_up <- all_diff_genes_df %>%
  dplyr::filter(hgnc_symbol %in% SIGNATURE_GENES,
                !is.na(padj), padj < PADJ_THRES, log2FoldChange > 0) %>%
  nrow()
sig_sig <- all_diff_genes_df %>%
  dplyr::filter(hgnc_symbol %in% SIGNATURE_GENES,
                !is.na(padj), padj < PADJ_THRES) %>%
  nrow()

target_present <- intersect(TARGET_GENES, all_diff_genes_df$hgnc_symbol)
target_sig <- target_gene_list %>%
  dplyr::filter(!is.na(padj), padj < PADJ_THRES) %>%
  dplyr::pull(hgnc_symbol)

# --- DE counts ---
n_tested <- nrow(all_diff_genes_df)
n_sig    <- sum(all_diff_genes_df$padj < PADJ_THRES, na.rm = TRUE)
n_up     <- sum(all_diff_genes_df$padj < PADJ_THRES &
                all_diff_genes_df$log2FoldChange > 0, na.rm = TRUE)
n_down   <- sum(all_diff_genes_df$padj < PADJ_THRES &
                all_diff_genes_df$log2FoldChange < 0, na.rm = TRUE)

# --- GO (BP chunk above) ---
n_go_bp <- if (exists("go_results") && !is.null(go_results)) nrow(as.data.frame(go_results)) else NA_integer_

summary_text <- sprintf(
"%s Differential Expression Report
=========================================
Date: %s

Configuration:
  Condition:        %s   (%s vs %s, reference = %s)
  Batches:          %s
  Count matrix:     %s
  Batch handling:   %s
  Design formula:   %s
  padj threshold:   %.3f
  Volcano labels:   %d
  enrichGO cutoff:  p < %.3f   (max %d bars)

Cohort:
  Total samples:       %d
  %s = %s (numerator): %d
  %s = %s (reference):  %d

Feature space:
  Genes in raw count table:     %d
  Protein-coding genes kept:    %d
  Genes tested by DESeq2:       %d

DESeq2 Results (%s: %s vs %s):
  Genes with padj < %.3f:       %d
  Up in %s:                     %d
  Down in %s:                   %d

Signature Genes (%d in panel):
  Detected in data:             %d
  Significant (padj < %.3f):    %d
  Up in %s (log2FC > 0):        %d

Target Genes (%d in panel):
  Detected in data:             %d
  Significant (padj < %.3f):    %s

Enrichment:
  enrichGO BP terms (p < %.3f): %s
  MSigDB categories run:        %s

Interpretation:
  %s

Figures (%s):
  00_meansd_before_vst.png      -> mean-SD, log2 normalized counts
  01_meansd_after_vst.png       -> mean-SD, after VST
  02_sample_distance_heatmap.png-> sample-sample Euclidean distances
  03_variable_gene_heatmap.png  -> top 50 most variable genes
  04_pca_condition.png          -> PCA coloured by %s
  05_pca_batch.png              -> PCA coloured by batch
  06_ma_plot.png                -> mean expression vs log2FC
  07_target_gene_counts.png     -> Normalized counts, target genes
  08_sig_gene_counts.png        -> Normalized counts, signature genes (+ CD200, KIT)
  09_top_gene_counts_1_4.png    -> Normalized counts, top DE genes 1-4
  10_top_gene_counts_5_8.png    -> Normalized counts, top DE genes 5-8
  11_volcano.png                -> log2FC vs -log10(padj)
  12-14_enrichgo_<onto>.png     -> enrichGO barplots (BP, CC, MF)
  15-26_msigdb_<cat>_*.png      -> fgsea table and enrichment curves per category

Tables (%s):
  all_diff_genes_df.csv        -> All genes tested (padj, log2FC, symbol)
  all_diff_genes.csv           -> All genes, NA-dropped, with significance flag
  diff_genes_significant.csv   -> Genes with padj < %.3f
  go_df_<onto>.csv             -> enrichGO results per ontology
  msigdb_results_<cat>_test.csv -> fgsea results per MSigDB category

Rendered report:
  %s
",
  toupper(CONDITION),
  format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z", tz = "America/New_York", usetz = TRUE),
  CONDITION, CONDITION_A, CONDITION_B, CONDITION_B,
  paste(BATCHES, collapse = ", "),
  basename(COUNT_TABLE_FILE),
  if (USE_ADJUSTED_COUNTS)
    "ComBat-seq adjusted upstream (group = condition); not in design"
  else
    "single batch; no adjustment",
  paste(deparse(DESIGN_FORMULA), collapse = ""),
  PADJ_THRES,
  NLABEL,
  PVALUECUTOFF, ENRICH_BARCOUNT,
  n_samples,
  CONDITION, CONDITION_A, n_a,
  CONDITION, CONDITION_B, n_b,
  nrow(count_table), nrow(data), n_tested,
  CONDITION, CONDITION_A, CONDITION_B,
  PADJ_THRES, n_sig,
  CONDITION_A, n_up,
  CONDITION_A, n_down,
  length(SIGNATURE_GENES),
  length(sig_present),
  PADJ_THRES, sig_sig,
  CONDITION_A, sig_up,
  length(TARGET_GENES),
  length(target_present),
  PADJ_THRES,
  if (length(target_sig)) paste(target_sig, collapse = ", ") else "none",
  PVALUECUTOFF,
  if (is.na(n_go_bp)) "not run" else as.character(n_go_bp),
  "H, C4, C5, C6, C7, C8",
  if (n_sig > 0)
    sprintf("%d genes differentially expressed at padj < %.3f; %d of %d signature genes reach significance.",
            n_sig, PADJ_THRES, sig_sig, length(sig_present))
  else
    sprintf("No genes reach padj < %.3f -- check group sizes (%d vs %d), batch confounding, and the PCA plot.",
            PADJ_THRES, n_a, n_b),
  FIG_DIR, CONDITION,
  TAB_DIR,
  PADJ_THRES,
  new_file_name
)

writeLines(summary_text, file.path(output_dir, "Z_results_summary.txt"))
message("\n=== SUMMARY ===")
message(summary_text)
message("\nAnalysis complete. Outputs saved to: ", normalizePath(output_dir))
```
